---
title: "world-population-over-300-years"
author: "Tanvi Hasija"
date: "2023-02-26"
output: html_document
knit: (
  function(input_file, encoding) {
    out_dir <- 'docs';
    rmarkdown::render(input_file,
      encoding=encoding,
      output_file=file.path(dirname(input_file), out_dir, 'index.html')
    )
  }
 )
---

### Dataset

We will use Gapminder Population Data version 7. It can be downloaded from [here](https://gapm.io/dl_popv7). Download this MS Excel file and save it in the `data` directory of this repo as `population_data.xlsx`.

### R Packages

#### Download the required packages.

Note: Run these commands manually in the R console

    install.packages("rnaturalearth")
    install.packages("stringr")
    install.packages("readxl")
    install.packages("dplyr")
    install.packages("ggplot2")
    install.packages("plotly")

#### Attach these packages

```{r}
library(rnaturalearth)
library(stringr)
library(readxl)
library(dplyr)
library(scales)
library(ggplot2)
library(plotly)
```

### Capturing and cleaning the data

#### Capture the World Map

```{r}
world_map = ne_countries(returnclass = "sf")
```

#### Read the dowloaded Population Data Xlsx and understand its structure

Note: the required data is in the 4th sheet of the MS Excel file.

```{r}
world_pop = read_xlsx("data/population_data.xlsx", sheet = 4)
names(world_pop)
```

#### Check for the differences in country names in the Population data and the world map data (for eg "United States" vs "USA")

```{r}
setdiff(world_map$sovereignt, world_pop$name)
```

#### Fix the anomallies and rename the column in the Population data to match with the column name in the world map data

```{r}
world_pop = world_pop %>% 
  mutate(sovereignt = name) %>% 
  mutate(sovereignt = replace(sovereignt, sovereignt == "Tanzania", "United Republic of Tanzania")) %>% 
  mutate(sovereignt = replace(sovereignt, sovereignt == "United States", "United States of America")) %>% 
  mutate(sovereignt = replace(sovereignt, sovereignt == "Congo, Dem. Rep.", "Democratic Republic of the Congo")) %>% 
  mutate(sovereignt = replace(sovereignt, sovereignt == "Bahamas", "The Bahamas")) %>% 
  mutate(sovereignt = replace(sovereignt, sovereignt == "Serbia", "Republic of Serbia")) %>% 
  mutate(sovereignt = replace(sovereignt, sovereignt == "Macedonia, FYR", "Macedonia")) %>% 
  mutate(sovereignt = replace(sovereignt, sovereignt == "Slovak Republic", "Slovakia")) %>% 
  mutate(sovereignt = replace(sovereignt, sovereignt == "Congo, Rep.", "Republic of Congo")) %>% 
  mutate(sovereignt = replace(sovereignt, sovereignt == "Kyrgyz Republic", "Kyrgyzstan")) %>% 
  mutate(sovereignt = replace(sovereignt, sovereignt == "Lao", "Laos")) %>% 
  mutate(sovereignt = replace(sovereignt, sovereignt == "Cote d'Ivoire", "Ivory Coast")) %>% 
  mutate(sovereignt = replace(sovereignt, sovereignt == "Timor-Leste", "East Timor")) %>% 
  mutate(sovereignt = replace(sovereignt, sovereignt == "Guinea-Bissau", "Guinea Bissau"))
```

#### Preparing the data

Now we can join our two datasets, pick the columns we want for visualizing and remove missing values.

```{r}
world_data = inner_join(world_map, world_pop, by = "sovereignt") %>%
  select(geo, sovereignt, Population, time, geometry ) %>%
  na.omit()
head(world_data)
```

We are only visualizing the data every two years, so we will remove the rest of the data too.
```{r}
world_data = world_data %>% 
  filter(time %in% seq(1800, 2100, by = 20))
head(world_data)
```

### Plotting

```{r}
ggplotly(
  ggplot(
    world_data,
    aes(geometry = geometry, frame = time)
    ) +
    geom_sf(aes(fill= Population)) +
    theme_void() +
    ggtitle("World Population Prediction 1800 - 2100") +
    scale_fill_distiller(
      palette = "RdBu",
      label = number_format(scale=1e-6, suffix = "M"),
      limits = c(0, 2000000000)
      )
  )
```
